On exponential sums arising from the classical doubling method

### 2025-02-09 paper.tex
Update from overleaf

\begin{proposition}
	We have the identity
	$$\Phi_{\chi}\left(g\right) = q^{-\frac{k^2}{2}} \GaussSumScalar{\chi_{\GL_k}}{\fieldCharacter} \sum_{\substack{x, y \in \GL_k\left(\finiteField\right)\\
			y^{-1} x = -g}} \fieldCharacter\left(\trace x\right) \chi^{-1}\left(\det y\right) \fieldCharacter\left(\trace y\right).$$
\end{proposition}
\begin{proof}
	We write $$\Phi_{\chi}\left(g\right) = \frac{1}{\sizeof{\squareMatrix_k\left(\finiteField\right)}}\sum_{X \in \squareMatrix_k\left(\finiteField\right)} \sum_{h \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace \left(X\left(g-\IdentityMatrix{k}-h\right)\right)\right) \chi\left(\det h\right).$$
	We claim that for a singular matrix $X \in \squareMatrix_k\left(\finiteField\right)$ the inner sum vanishes. To show this, write $$X = h_1 \begin{pmatrix}
		\IdentityMatrix{k-r}\\
		& 0_r
	\end{pmatrix} h_2,$$
	where $1 \le r \le k$ and $h_1, h_2 \in \GL_k\left(\finiteField\right)$. Then
	$$\fieldCharacter\left(\trace Xh\right) = \fieldCharacter\left(\trace\left( \begin{pmatrix}
		\IdentityMatrix{k-r}\\
		& 0_r
	\end{pmatrix} h_2 h h_1\right)\right).$$
	Changing variables $h \mapsto h_2^{-1} h h_1^{-1},$
	we obtain the inner sum
	$$\sum_{h \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace \begin{pmatrix}
		\IdentityMatrix{k-r}\\
		& 0_r
	\end{pmatrix} h \right) \chi\left(\det h\right),$$
	which equals
	$$\frac{1}{\sizeof{\multiplicativegroup{\finiteField}}} \sum_{a \in \multiplicativegroup{\finiteField}} \sum_{h \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace \begin{pmatrix}
		\IdentityMatrix{k-r}\\
		& 0_r
	\end{pmatrix} \begin{pmatrix}
		\IdentityMatrix{k-1}\\
		& a
	\end{pmatrix} h \right) \chi\left(\det h\right).$$
	Changing variables $h \mapsto \left(\begin{smallmatrix}
		\IdentityMatrix{k-1}\\
		& a^{-1}
	\end{smallmatrix}\right) h$, we get the inner sum $$\sum_{a \in \multiplicativegroup{\finiteField}} \chi^{-1}\left(a\right),$$
	which vanishes because $\chi \ne 1$.
	Hence, we have
	$$\Phi_{\chi}\left(g\right) = \frac{1}{\sizeof{\squareMatrix_k\left(\finiteField\right)}} \sum_{X, h \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace \left(X\left(g-\IdentityMatrix{k}-h\right)\right)\right) \chi\left(\det h\right).$$
	Changing variables $h \mapsto X^{-1} h$, we get
	$$\Phi_{\chi}\left(g\right) = \frac{1}{\sizeof{\squareMatrix_k\left(\finiteField\right)}} \sum_{X, h \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace \left(Xg\right)\right) \fieldCharacter\left(-\trace h\right) \chi\left(\det h\right) \chi^{-1}\left(\det X\right) \fieldCharacter\left(-\trace X\right).$$
	Changing variables again $h \mapsto -h$ and setting $Xg = x$ and $-X = y$, we get the equality 
	$$\Phi_{\chi}\left(g\right) = q^{-\frac{k^2}{2}} \GaussSumScalar{\chi_{\GL_k}}{\fieldCharacter} \sum_{\substack{x, y \in \GL_k\left(\finiteField\right)\\
			y^{-1} x = -g}} \fieldCharacter\left(\trace x\right) \chi^{-1}\left(\det y\right) \fieldCharacter\left(\trace y\right).$$
\end{proof}

@@

\subsubsection{General linear groups computation}
Suppose that $F \colon \GL_k\left(\finiteField\right) \to \cComplex$ is a class function. If $\chi \colon \multiplicativegroup{\finiteField} \to \multiplicativegroup{\cComplex}$ is a non-trivial character, we define $$\dblVirtualGaussSumScalar{F}{\chi} = q^{-\frac{k^2}{2}} \sum_{g \in \GL_k\left(\finiteField\right)} \Phi_{\chi}\left(g\right) F\left(g\right).$$
It is clear that for an irreducible representation $\tau$ of $\GL_k\left(\finiteField\right)$, we have $$\dblVirtualGaussSumScalar{\trace \tau}{\chi} = \dblGaussSumScalar{\tau}{\chi} \cdot \dim \tau.$$
Since the assignment $g \mapsto \Phi_{\chi}\left(g\right)$ depends only on the semisimple part of $g$, by \cite[Theorem in Section 1.2]{SaitoShinoda2000} we have that if $T \cong \multiplicativegroup{\finiteFieldExtension{k}}$ and if $\theta \colon \multiplicativegroup{\finiteFieldExtension{k}} \to \multiplicativegroup{\cComplex}$ is a character then
$$\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} = \frac{\grpIndex{\GL_k\left(\finiteField\right)}{T}}{q^{\frac{k^2}{2}}} \sum_{\substack{t \in \multiplicativegroup{\finiteFieldExtension{k}}\\
t \ne 1}} \theta\left(t\right) \chi\left(\FieldNorm{k}{1}\left(t-1\right)\right).$$
Using \Cref{thm:gln-doubling-gauss-sum-in-terms-of-kondo} and the fact that Kondo's Gauss sum is the same for every subrepresentation of $R_{T, \theta}$, we have the equality $$\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} = \left(-1\right)^k \theta\left(-1\right) q^{-\frac{3k}{2}} R_{T, \theta}\left(1\right) \tau\left(\chi, \fieldCharacter\right)^k \tau\left(\theta, \fieldCharacter\right) \tau\left(\theta^{-1} \times \chi^{-1}, \fieldCharacter\right).$$

@@

Hence we get
$$\sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{k}}\\
		x \ne 1}} \theta \left(x\right) \chi\left(\FieldNorm{k}{1}\left(x - 1\right)\right) \chi\left(\FieldNorm{k}{1}\left(x^{-1} - 1\right)\right) = \frac{q^{\frac{k^2}{2}}}{\grpIndex{\GL_k\left(\finiteField\right)}{T}} \dblVirtualGaussSumScalar{R_{T, \theta_{\chi^{-1}}}}{\chi^2},$$
which by the results of the previous section equals
%$$\frac{q^{\frac{k^2}{2}}}{\grpIndex{\GL_k\left(\finiteField\right)}{T}} \left(-1\right)^k \theta\left(-1\right) q^{-\frac{3k}{2}} R_{T, \theta}\left(1\right) \tau\left(\chi^2, \fieldCharacter\right)^k \tau\left(\theta, \fieldCharacter\right) \tau\left(\theta^{-1} \times \chi^{-1}, \fieldCharacter\right). $$
%$$\frac{q^{\frac{k^2}{2}}}{\grpIndex{\GL_k\left(\finiteField\right)}{T}} \left(-1\right)^k \theta\left(-1\right) q^{-\frac{3k}{2}} \left(-1\right)^{k-1} \frac{\grpIndex{\GL_k\left(\finiteField\right)}{T}}{q^{\frac{k^2 - k}{2}}} \tau\left(\chi^2, \fieldCharacter\right)^k \tau\left(\theta, \fieldCharacter\right) \tau\left(\theta^{-1} \times \chi^{-1}, \fieldCharacter\right).$$
%$$R_{T,\theta}\left(1\right) = \left(-1\right)^{k-1} \frac{\grpIndex{\GL_k\left(\finiteField\right)}{T}}{q^{\frac{k^2 - k}{2}}}$$
$$-\theta\left(-1\right) q^{-k} \tau\left(\chi^2, \fieldCharacter\right)^k \tau\left(\theta \times \chi^{-1}, \fieldCharacter\right) \tau\left(\theta^{-1} \times \chi^{-1}, \fieldCharacter\right).$$

@@

In this appendix, we evaluate the following exponential sum. Let $\chi \colon \multiplicativegroup{\finiteField} \to \multiplicativegroup{\cComplex}$ be a non-trivial character. We compute $$\sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{2}}\\
	\trace_{\finiteFieldExtension{2} \slash \finiteField}\left(x\right) \ne 0}} \chi\left(\FieldNorm{2}{1}\left(\frac{x}{\trace_{\finiteFieldExtension{2} \colon \finiteField} x}\right) \right).$$
As usual, we rewrite this sum as
$$\frac{1}{q} \sum_{z \in \finiteField} \sum_{y \in \multiplicativegroup{\finiteField}} \sum_{x \in \multiplicativegroup{\finiteFieldExtension{2}}} \chi\left(\FieldNorm{2}{1}\left(xy\right)\right) \fieldCharacter\left(z\left(y \trace_{\finiteFieldExtension{2} \slash \finiteField} x - 1\right)\right).$$
It is clear that if $z = 0$, then the sum over $x$ will vanish. Hence, we may reduce the sum to a sum over $z \in \multiplicativegroup{\finiteField}$. By replacing $y$ with $z^{-1} y$ and then replacing $z$ with $-z$, we arrive at the sum
$$\frac{1}{q} \sum_{y \in \multiplicativegroup{\finiteField}} \sum_{x \in \multiplicativegroup{\finiteFieldExtension{2}}} \chi\left(\FieldNorm{2}{1}\left(xy\right)\right) \fieldCharacter\left(y \trace_{\finiteFieldExtension{2} \slash \finiteField} x\right) \sum_{z \in \multiplicativegroup{\finiteField}} \chi^{-2}\left(z\right)\fieldCharacter\left(z\right).$$
Replacing $x$ with $y^{-1}x$, and using the Hasse--Davenport relation, we arrive at the sum
$$q^{\frac{1}{2}} \left(q-1\right) \tau\left(\chi, \fieldCharacter\right)^2 \tau\left(\chi^{-2}, \fieldCharacter\right).$$
\bibliographystyle{abbrv}
\bibliography{references}
\end{document}

>>>>

\begin{proposition}
	We have the identity
	$$\Phi_{\chi}\left(g\right) = q^{-\frac{k^2}{2}} \GaussSumScalar{\chi_{\GL_k}}{\fieldCharacter} \sum_{\substack{x, y \in \GL_k\left(\finiteField\right)\\
			y^{-1} x = -g}} \fieldCharacter\left(\trace x\right) \chi^{-1}\left(\det y\right) \fieldCharacter\left(\trace y\right).$$
\end{proposition}
\begin{proof}
	We write $$\Phi_{\chi}\left(g\right) = \frac{1}{\sizeof{\squareMatrix_k\left(\finiteField\right)}}\sum_{X \in \squareMatrix_k\left(\finiteField\right)} \sum_{h \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace \left(X\left(g-\IdentityMatrix{k}-h\right)\right)\right) \chi\left(\det h\right).$$
	We claim that for a singular matrix $X \in \squareMatrix_k\left(\finiteField\right)$ the inner sum vanishes. To show this, write $$X = h_1 \begin{pmatrix}
		\IdentityMatrix{k-r}\\
		& 0_r
	\end{pmatrix} h_2,$$
	where $1 \le r \le k$ and $h_1, h_2 \in \GL_k\left(\finiteField\right)$. Then
	$$\fieldCharacter\left(\trace Xh\right) = \fieldCharacter\left(\trace\left( \begin{pmatrix}
		\IdentityMatrix{k-r}\\
		& 0_r
	\end{pmatrix} h_2 h h_1\right)\right).$$
	Changing variables $h \mapsto h_2^{-1} h h_1^{-1},$
	we obtain the inner sum
	$$\sum_{h \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace \begin{pmatrix}
		\IdentityMatrix{k-r}\\
		& 0_r
	\end{pmatrix} h \right) \chi\left(\det h\right),$$
	which equals
	$$\frac{1}{\sizeof{\multiplicativegroup{\finiteField}}} \sum_{a \in \multiplicativegroup{\finiteField}} \sum_{h \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace \begin{pmatrix}
		\IdentityMatrix{k-r}\\
		& 0_r
	\end{pmatrix} \begin{pmatrix}
		\IdentityMatrix{k-1}\\
		& a
	\end{pmatrix} h \right) \chi\left(\det h\right).$$
	Changing variables $h \mapsto \left(\begin{smallmatrix}
		\IdentityMatrix{k-1}\\
		& a^{-1}
	\end{smallmatrix}\right) h$, we get the inner sum $$\sum_{a \in \multiplicativegroup{\finiteField}} \chi^{-1}\left(a\right),$$
	which vanishes because $\chi \ne 1$.
	Hence, we have
	$$\Phi_{\chi}\left(g\right) = \frac{1}{\sizeof{\squareMatrix_k\left(\finiteField\right)}} \sum_{X, h \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace \left(X\left(g-\IdentityMatrix{k}-h\right)\right)\right) \chi\left(\det h\right).$$
	Changing variables $h \mapsto X^{-1} h$, we get
	$$\Phi_{\chi}\left(g\right) = \frac{1}{\sizeof{\squareMatrix_k\left(\finiteField\right)}} \sum_{X, h \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace \left(Xg\right)\right) \fieldCharacter\left(-\trace h\right) \chi\left(\det h\right) \chi^{-1}\left(\det X\right) \fieldCharacter\left(-\trace X\right).$$
	Changing variables again $h \mapsto -h$ and setting $Xg = x$ and $-X = y$, we get  
	$$\Phi_{\chi}\left(g\right) = q^{-\frac{k^2}{2}} \GaussSumScalar{\chi_{\GL_k}}{\fieldCharacter} \sum_{\substack{x, y \in \GL_k\left(\finiteField\right)\\
			y^{-1} x = -g}} \fieldCharacter\left(\trace x\right) \chi^{-1}\left(\det y\right) \fieldCharacter\left(\trace y\right).$$
\end{proof}

@@

\subsubsection{General linear groups computation}
Suppose that $F \colon \GL_k\left(\finiteField\right) \to \cComplex$ is a class function. If $\chi \colon \multiplicativegroup{\finiteField} \to \multiplicativegroup{\cComplex}$ is a non-trivial character, we define $$\dblVirtualGaussSumScalar{F}{\chi} = q^{-\frac{k^2}{2}} \sum_{g \in \GL_k\left(\finiteField\right)} \Phi_{\chi}\left(g\right) F\left(g\right).$$
It is clear that for an irreducible representation $\tau$ of $\GL_k\left(\finiteField\right)$, we have $$\dblVirtualGaussSumScalar{\trace \tau}{\chi} = \dblGaussSumScalar{\tau}{\chi} \cdot \dim \tau.$$
Since the assignment $g \mapsto \Phi_{\chi}\left(g\right)$ depends only on the semisimple part of $g$, by \cite[Theorem in Section 1.2]{SaitoShinoda2000} we have that if $T \cong \multiplicativegroup{\finiteFieldExtension{k}}$ and if $\theta \colon \multiplicativegroup{\finiteFieldExtension{k}} \to \multiplicativegroup{\cComplex}$ is a character then
$$\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} = \frac{\grpIndex{\GL_k\left(\finiteField\right)}{T}}{q^{\frac{k^2}{2}}} \sum_{\substack{t \in \multiplicativegroup{\finiteFieldExtension{k}}\\
t \ne 1}} \theta\left(t\right) \chi\left(\FieldNorm{k}{1}\left(t-1\right)\right).$$
Using \Cref{thm:gln-doubling-gauss-sum-in-terms-of-kondo} and the fact that Kondo's Gauss sum is the same for every subrepresentation of $R_{T, \theta}$, we have the equality $$\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} = \left(-1\right)^k \theta\left(-1\right) R_{T, \theta}\left(1\right) \tau\left(\chi, \fieldCharacter\right)^k \tau\left(\theta, \fieldCharacter\right) \tau\left(\theta^{-1} \times \chi^{-1}, \fieldCharacter\right).$$

@@

Hence we get
$$\sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{k}}\\
		x \ne 1}} \theta \left(x\right) \chi\left(\FieldNorm{k}{1}\left(x - 1\right)\right) \chi\left(\FieldNorm{k}{1}\left(x^{-1} - 1\right)\right) = \frac{q^{\frac{k^2}{2}}}{\grpIndex{\GL_k\left(\finiteField\right)}{T}} \dblVirtualGaussSumScalar{R_{T, \theta_{\chi^{-1}}}}{\chi^2},$$
which by the results of the previous section equals
$$-\theta\left(-1\right) q^{\frac{k}{2}} \tau\left(\chi^2, \fieldCharacter\right)^k \tau\left(\theta \times \chi^{-1}, \fieldCharacter\right) \tau\left(\theta^{-1} \times \chi^{-1}, \fieldCharacter\right).$$

@@

In this appendix, we evaluate the following exponential sum. Let $\chi \colon \multiplicativegroup{\finiteField} \to \multiplicativegroup{\cComplex}$ be a non-trivial character. We compute $$\sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{2}}\\
	\trace_{\finiteFieldExtension{2} \slash \finiteField}\left(x\right) \ne 0}} \chi\left(\FieldNorm{2}{1}\left(\frac{x}{\trace_{\finiteFieldExtension{2} \slash \finiteField} x}\right) \right).$$
As usual, we rewrite this sum as
$$\frac{1}{q} \sum_{z \in \finiteField} \sum_{y \in \multiplicativegroup{\finiteField}} \sum_{x \in \multiplicativegroup{\finiteFieldExtension{2}}} \chi\left(\FieldNorm{2}{1}\left(xy\right)\right) \fieldCharacter\left(z\left(y \trace_{\finiteFieldExtension{2} \slash \finiteField} x - 1\right)\right).$$
It is clear that if $z = 0$, then the sum over $x$ will vanish. Hence, we may reduce the sum to a sum over $z \in \multiplicativegroup{\finiteField}$. By replacing $y$ with $z^{-1} y$ and then replacing $z$ with $-z$, we arrive at the sum
$$\frac{1}{q} \sum_{y \in \multiplicativegroup{\finiteField}} \sum_{x \in \multiplicativegroup{\finiteFieldExtension{2}}} \chi\left(\FieldNorm{2}{1}\left(xy\right)\right) \fieldCharacter\left(y \trace_{\finiteFieldExtension{2} \slash \finiteField} x\right) \sum_{z \in \multiplicativegroup{\finiteField}} \chi^{-2}\left(z\right)\fieldCharacter\left(z\right).$$
Replacing $x$ with $y^{-1}x$, and using the Hasse--Davenport relation, we arrive at the sum
$$q^{\frac{1}{2}} \left(q-1\right) \tau\left(\chi, \fieldCharacter\right)^2 \tau\left(\chi^{-2}, \fieldCharacter\right).$$
\bibliographystyle{abbrv}
\bibliography{references}
\end{document}
